---
title: "Exercise Title Goes Here"
author: "Your Name Goes Here"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

<!-- These are global options  
       set root directory here 
       NOTE mine is active; yours is commented out
       change the path to reflect your CPU,
       then comment out or delete mine
--> 
```{r global_options}
knitr::opts_knit$set(root.dir = "D:/OneDrive - University of Vermont/Classes/Spring2022/sdhmR/sdhmR-V2022.1")
#knitr::opts_knit$set(root.dir = "~/sdhmR-V2022.1")
knitr::opts_chunk$set(warning=FALSE,
                      error=TRUE,
                      message=FALSE)
#Setting Paths
path.root <- "D:/OneDrive - University of Vermont/Classes/Spring2022/sdhmR/sdhmR-V2022.1"
path.ex <- paste(path.root, "/data/exercise/traindat", sep = "") #Path to mod 2
path.preds <- paste(path.root, '/data/exercise/preds', sep ='')
path.figs <- paste(path.root, "/powerpoints/figures", sep = "") #path to save figs
path.gis <- paste(path.root, "/data/gis_layers", sep = "") #path to gis files

# some libraries
library(raster)	# FXNS: extent, raster, res, rasterize, values, writeRaster, 
#       cellFromXY, extract
library(sf)    	# FXNS: st_as_sf, st_as_sfc, st_bbox, st_convex_hull, st_union
#       st_buffer, st_read, st_write, st_intersection, st_geometry,
#       st_drop_geometry, st_coordinates, st_sf
library(terra)	# FXNS: as.data.frame, rast, vect
library(sp)

# source course CRSs
setwd(path.root)
source("r_code/prjs4_sdhmR.r") # source course CRSs
ls(pattern = "prj.") # should be 5 prj.* objects
```

---

## This exercise links to Module 2.5

Submit the completed exercise to us at:

* t.edwards@usu.edu 
* eric_tsakiris@fws.gov  

---

## Context

The data are of the *Pinus edulis* Common piñon, the species you dealt with in Exercise #3 -- #4.  This exercise relies on the pseudo--absence values you generated in exercise #4, and then subsequently added to the data presence values dataframe.  Hopefully you saved those combined presence:absence dataframe, or regrettably you will need to return to exercise #4 and re--build that dataframe.

The goal is to use the dataframe consisting of the presence:absence [X,Y]'s from exercise #4, and "punch" those [X,Y]'s through a stack of environmental predictor variables.  The extracted values are to be added to the exercise #4 dataframe, creating a new dataframe consisting of the: 

* FNETID ID, 
* the [X,Y]'s in the correct CRS, 
* the presence:absence coded as [0=absence, 1=presence], and 
* the predictor values associated with [X,Y] in the dataframe.

Remember, you will have been assigned data for *edulis* having one the following four discrete labels:

* **Seedlings**:  these data indicate spatial locations where seedlings of the tree have been found
* **Mortality**: these data indicate spatial locations where mortality has been observed
* **Persistence**: these data have no observed mortality or seedling at given spatial locations
* **Range**: the total dataset, including all spatial locations of seedlings, mortality, and persistence

---

## The Data

* Common piñon *Pinus edulis* sppcode=106
* The pseudo-absence data: Wherever stored from Homework #4
* Predictors: **`~\sdhmR-V2022.1\data\exercise\preds`** 
  * There are a total of 16 predictor variables the directory above
  * The variables are **`.img`** files
* Data files:
  * **`spp106pr_MORT.csv`**, **`spp106pr_PERS.csv`**, **`spp106pr_RANG.csv`**, **`spp106pr_SEED.csv`**

## Global Options



---

## Question #1

* Import the predictor GIS variables into R

```{r}
setwd(path.preds)
etpt_5<- raster('etpt_5.img')
etpt_6 <- raster('etpt_6.img')
etpt_sprin <- raster('etpt_sprin.img')
exp1nrm <- raster('exp1nrm.img')
exp3nrm <- raster('exp3nrm.img')
exp5nrm <- raster('exp5nrm.img')
mind_yr_av <- raster('mind_yr_av.img')
prad_sw_di <- raster('prad_sw_di.img')
prec_w_hal <- raster('prec_w_hal.img')
prec_winte <- raster('prec_winte.img')
rough_1k <- raster('rough_1k.img')
tave_s_hal <- raster('tave_s_hal.img')
tave_sprin <- raster('tave_sprin.img')
tmax_summe <- raster('tmax_summe.img')
topos <- raster('topos.img')

#Check projections
projection(etpt_5)
projection(etpt_6)
projection(etpt_sprin)
projection(exp3nrm)
projection(exp5nrm)
projection(mind_yr_av)
projection(prad_sw_di)
projection(prad_sw_di)
projection(prec_w_hal)
projection(prec_winte)
projection(rough_1k)
projection(tave_s_hal)
projection(tave_sprin)
projection(tmax_summe)
projection(topos)

#compare raster resolution
res(etpt_5)
res(etpt_6)
res(etpt_sprin)
res(exp3nrm)
res(exp5nrm)
res(mind_yr_av)
res(prad_sw_di)
res(prad_sw_di)
res(prec_w_hal)
res(prec_winte)
res(rough_1k)
res(tave_s_hal)
res(tave_sprin)
res(tmax_summe)
res(topos)
#all resolutions look the same, so no need to convert

```


---

## Question #2

* Extract predictor variables from all 16 **`.img`** files using the presence:absence [X,Y]'s

```{r}
setwd(path.ex)
pres.abs.pers <- get(load("pers.PPsA.RData"))
load('pers.bufR.RData')

setwd(path.preds)
preds.list <- list.files(pattern = ".img$") # list of .img files; $ strips extra
preds.list # examine

# loop for extracting topo data w/raster stack
layers <- {} # initialize (empty) list of raster layers
for (i in 1:length(preds.list)) {
  r1 <- crop(raster(preds.list[i]), pers.bufR) # crop pred var raster to buffer
  names(r1) <- strsplit(preds.list[i], "_wgs.img") # assign name to raster
  layers <- c(layers, r1) # add raster to layer of rasters
}
layers # examine raster stack; return is a list of raster layers

# build the raster stack
pers.topoDOM <- stack(layers) # create a raster stack
pers.topoDOM # examine stack

# extract from stack 
t1 <- extract(pers.topoDOM, pres.abs.pers[, c("tr.wgs_x", "tr.wgs_y")]) # extract values from raster stack

```

---

## Question #3

* Create a new dataframe by binding the extracted predictor variable values to the dataframe from exercise #4.

```{r}

head(t1, 2) # examine extracted matrix
pers.trTOPO <- cbind(pres.abs.pers, t1) # bind to train dataframe
head(pers.trTOPO, 2) # examine training data frame

```

---

## Question #4

* Save your data as R objects:
  * Dataframe;
  * Point shapefile with geometry in R
  * Export as a point shapefile in ESRI format
* Save these R objects in a **`.RData`** file

These data will be used in the next exercise #6.

```{r}

setwd(path.ex)
write.csv(pers.trTOPO, file = "pers_trTOPO.csv", row.names = F) # save .csv
save(pers.trTOPO, file = "pers.trTOPO.RData") # save .RData
save(pers.topoDOM, file = "pers.topoDOM.RData") # save .RData

```

---

## The End

---
