---
title: "Exercise Title Goes Here"
author: "Your Name Goes Here"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

<!-- These are global options  
       set root directory here 
       NOTE mine is active; yours is commented out
       change the path to reflect your CPU,
       then comment out or delete mine
--> 
```{r global_options, include=FALSE}
knitr::opts_knit$set(root.dir = "D:/OneDrive - University of Vermont/Classes/Spring2022/sdhmR/sdhmR-V2022.1") #Reed Path
#knitr::opts_knit$set(root.dir ="C:/Users/14842/Documents/SDHM/sdhmR-V2022.1") #Lindsey Path
#knitr::opts_knit$set(root.dir = "~/sdhmR-V2022.1")
knitr::opts_chunk$set(warning=FALSE,error=TRUE,message=FALSE)
```

---

## This exercise links to Module 7

Submit the completed exercise to us at:

* t.edwards@usu.edu 
* eric_tsakiris@fws.gov  

---

## Context

The data are of the *Pinus edulis* Common piñon, the species you dealt with in Exercise #3 -- #5.  This exercise relies on the dataframe you generated in exercise #6.  Hopefully you saved those combined presence:absence dataframe, with the reduced variables you begin SDHM model construction with, or regrettably you will need to return to exercise #6 and re--build that dataframe.

The goal is to: 

* Build a dataframe of [X,Y] points selected for a field sample campaign
  * (You need only show the first 10 lines of the dataframe)
* Build ensemble concordance map products of 3, 5, and union of all models 
* Build tables of frequencies: 
  * Presences (ie, sum the cells of classified presence for the 3, 5, and union ensembles) 
  * Concordances (remember, this is a measure of "how much !!) 

Remember, you will have been assigned data for *edulis* having one the following four discrete labels:

* **Seedlings**:  these data indicate spatial locations where seedlings of the tree have been found
* **Mortality**: these data indicate spatial locations where mortality has been observed
* **Persistence**: these data have no observed mortality or seedling at given spatial locations
* **Range**: the total dataset, including all spatial locations of seedlings, mortality, and persistence

---

## There Are Different Objectives for Each of the Group Labels -- Pay Attention Here !!

* **RANG** => SRS based on bounding box, with an $n$ = 250
* **SEED** => SRS based on species modelling, with an $n$ = 250
* **PERS**=> SRS based on concordance strata, within bounding box, with $n$'s of:
  * N[0,1,2,3,4,5] = c(25, 45, 45, 45, 45, 45)
* **MORT** => SRS based on concordance strata, within modeling domain, with $n$'s of:
  * N[0,1,2,3,4,5] = c(25, 45, 45, 45, 45, 45)

---

## The Data

* Load:
  * The previously constructed SDHM models, for your group label, and 
  * The (i) prediction; and (ii) classified maps

If you did **NOT** save these you're in a bit of trouble ...

---

## Question #1

* Construct maps showing the sample points, by your respective group and its unique objective


```{r Initialize}
#Initialize -----------------------------------------
path.root <- "D:/OneDrive - University of Vermont/Classes/Spring2022/sdhmR/sdhmR-V2022.1" #Reed Laptop Path
#path.root <- "C:/Users/14842/Documents/SDHM/sdhmR-V2022.1" #Lindsey Path
path.ex <- paste(path.root, "/data/exercise/traindat", sep = "") #Path to mod 2
path.preds <- paste(path.root, '/data/exercise/preds', sep = '')
path.figs <- paste(path.root, "/powerpoints/figures", sep = "") #path to save figs
path.gis <- paste(path.root, "/data/gis_layers", sep = "") #path to gis files

# some libraries
library(raster)	# FXNS: extent, raster, res, rasterize, values, writeRaster, 
#       cellFromXY, extract
library(sf)    	# FXNS: st_as_sf, st_as_sfc, st_bbox, st_convex_hull, st_union
#       st_buffer, st_read, st_write, st_intersection, st_geometry,
#       st_drop_geometry, st_coordinates, st_sf
library(terra)	# FXNS: as.data.frame, rast, vect
library(sp)
library(dplyr)
library(tidyverse)
library(data.table)
library(modeest)
library(DAAG)
library(PresenceAbsence)
options("rgdal_show_exportToProj4_warnings"="none") # run this before library(rgdal)
library(rgdal)
library(dismo)
library(maptools)
library(randomForest)
library(gbm)
# source course CRSs
setwd(path.root)
getwd()
source("r_code/prjs4_sdhmR.r")
ls(pattern = "prj.") # should be 5 prj.* objects

set.seed(1234)

##########################################################
# Global Variables-------------------------------------
#load in prior data#---------------
setwd(path.ex)
load('tr_PERS.RData')
load('pres.bufSF.RDS')
load("pers.bufptR.img")
load('pers.PPsA.RData', verbose = TRUE)
#load in ensemble model
setwd(path.ex)
load('ensemble.dom.RData', verbose = T)
#load in ensemle stats
setwd(path.ex)
load('ensembleSTATS.RData')
#Read in Frames
setwd(path.gis)
states <- st_read(dsn = ".", layer = "na_states_wgs") # import shapefile

setwd(path.preds)
load('pers.topoDOM.RData')
pers.dom <- pers.topoDOM

setwd(path.ex)
load('pres.fnet.RData', verbose = TRUE)
class(tr_PERS)
head(tr_PERS)
drops <- c('wgs_xF','wgs_yF','cell.wgs_x','cell.wgs_y','exp5nrm','exp3nrm','rough_1k','prad_sw_di',
           'etpt_5','mind_yr_av','prec_winte','tave_sprin','UNIQUEID.x'
           ,'UNIQUEID.y')
tr_PERS <- tr_PERS[ ,!(names(tr_PERS) %in% drops)]
tr_PERS <- na.omit(tr_PERS)
dim(tr_PERS)
```

```{r Read in fnet. remove P/A points}
pres.fnetSP <- as_Spatial(pres.fnetSF)
fnet1 <- crop(rasterize(pres.fnetSP,pers.dom[[1]]),pers.bufptR)
fnet2 <- as.data.frame(fnet1)
head(fnet2)

#remove duplicate fnetids from p/a data

pers.PA <- subset(pers.PPsA, !duplicated(pers.PPsA[,1]))
head(pers.PA,2)

#remove spp P/A from fishet, fnet3 will be sample frame for field test
fnet3 <- as.data.frame(fnet2[!fnet2$FNETID %in% pers.PA,])
head(fnet3)
dim(fnet3)


#create final df for sample selection: FNETID and cell x&y
fnet.merge <- merge(fnet2,fnet3, by = c('FNETID','cell.wgs_x','cell.wgs_y'))
head(fnet.merge,2)
dim(fnet.merge)



fnet2SF <- st_as_sf(fnet2, coords = c('cell.wgs_x', 'cell.wgs_y'), crs = prj.wgs84)
plot(fnet2SF$geometry)
plot(pers.bufptR, add =T)
plot(pres.bufSF, add = T)
#it works! mostly#
```

```{r Build frame for extraction}
#####################################################################################
#Build sample frames for field sample extractions

sample.dom <- stack(prob.mean,probSTD.mean,clas.sum)
names(sample.dom) <- c('prob.mean','probSTD.mean','clas.sum')
sample.dom

# extract data from rasters & bind to sample dataframe
ext.1 <- raster::extract(x = sample.dom, y = fnet2[, 2:3], method = 'simple') # basic extract
ext.1 <- as.data.frame(ext.1)
fr.2sample <- cbind(fnet.merge, ext.1) # bind extracted values to sample DF
head(fr.2sample, 2) # examine
```

```{r Extract sample points + map}
#Simple random sample inluceding bbox
#srs draw from sample from
#set.seed(1234)
sample.size=250
srs1 <- fr.2sample[sample(1:nrow(fr.2sample),sample.size),]
head(srs1,2)
dim(srs1)

## giggleplot includes bbox:
pin.fig <- c(5.373, 3.623)  # fig size
par(pin = pin.fig)  # set fig par

#plot(fnet1, col = "white", legend = F)  # spp bbox
plot(pers.bufptR, col = "lightskyblue4", legend = F) # spp bbox
points(pers.PPsA$wgs_x, pers.PPsA$wgs_y, pch = 16, col= "snow4")
points(srs1$cell.wgs_x, srs1$cell.wgs_y, pch = 16, col = "tomato3") # add spp P/A points
legend("bottomleft", title = "SRS sample points",
       legend = c("Species domain", "Presence points", "SRS sample points"), bg = "white",
       cex = .75, bty = "0", inset = c(0.05, 0.05), pt.cex=1.5, pch = c(15, 15, 15),
       col = c("lightskyblue4", "snow4", "tomato3"))
```



---

## Question #2

* Export field sample points as a shapefile

```{r}
head(srs1,2) #examine dataframe for export

#build spatial dataset
pers.srs <- srs1[c('FNETID','cell.wgs_x','cell.wgs_y')]
names(pers.srs) <- c('FNETID','wgs84_x','wgs84_y')
head(pers.srs)
#convert to sf
pers.srsSF <- st_as_sf(pers.srs, coords = c('wgs84_x','wgs84_y'),
                       crs = prj.wgs84, remove =F) #remove=F to retain input x,y
head(pers.srsSF,2)

setwd(path.gis) # output path
st_write(pers.srsSF, dsn = ".", layer = "pers.srs", 
         driver = "ESRI Shapefile", delete_dsn = T) # output shapefile
```
---

## Question #3

* Save your data as R objects:
  * All field campaign maps as R spatial object
  * All field campaign maps as **`.img`** format
* Save these R objects in a **`.RData`** file as well

```{r}
pers.srsDF <- st_drop_geometry(pers.srsSF)
head(pers.srsDF,2)
save(pers.srsSF,pers.srsDF, file = 'ex13.RData')
savePlot(filename = "ex13fig01.pdf", type = "pdf")
```

---

## The End

---
