---
title: "Exercise Title Goes Here"
author: "Your Name Goes Here"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

<!-- These are global options  
       set root directory here 
       NOTE mine is active; yours is commented out
       change the path to reflect your CPU,
       then comment out or delete mine
--> 
```{r global_options, include=FALSE}
knitr::opts_knit$set(root.dir = "~/words/classes/sdmR_ALLversions/sdhmR-V2022.1")
#knitr::opts_knit$set(root.dir = "~/sdhmR-V2022.1")
knitr::opts_chunk$set(warning=FALSE,error=TRUE,message=FALSE)
```

---

## This exercise links to Module 7

Submit the completed exercise to us at:

* t.edwards@usu.edu 
* eric_tsakiris@fws.gov  

---

## Context

The data are of the *Pinus edulis* Common piñon, the species you dealt with in Exercise #3 -- #5.  This exercise relies on the dataframe you generated in exercise #6.  Hopefully you saved those combined presence:absence dataframe, with the reduced variables you begin SDHM model construction with, or regrettably you will need to return to exercise #6 and re--build that dataframe.

The goal is to: 

* Build a dataframe of [X,Y] points selected for a field sample campaign
  * (You need only show the first 10 lines of the dataframe)
* Build ensemble concordance map products of 3, 5, and union of all models 
* Build tables of frequencies: 
  * Presences (ie, sum the cells of classified presence for the 3, 5, and union ensembles) 
  * Concordances (remember, this is a measure of "how much !!) 

Remember, you will have been assigned data for *edulis* having one the following four discrete labels:

* **Seedlings**:  these data indicate spatial locations where seedlings of the tree have been found
* **Mortality**: these data indicate spatial locations where mortality has been observed
* **Persistence**: these data have no observed mortality or seedling at given spatial locations
* **Range**: the total dataset, including all spatial locations of seedlings, mortality, and persistence

---

## There Are Different Objectives for Each of the Group Labels -- Pay Attention Here !!

* **RANG** => SRS based on bounding box, with an $n$ = 250
* **SEED** => SRS based on species modelling, with an $n$ = 250
* **PERS**=> SRS based on concordance strata, within bounding box, with $n$'s of:
  * N[0,1,2,3,4,5] = c(25, 45, 45, 45, 45, 45)
* **MORT** => SRS based on concordance strata, within modeling domain, with $n$'s of:
  * N[0,1,2,3,4,5] = c(25, 45, 45, 45, 45, 45)

---

## The Data

* Load:
  * The previously constructed SDHM models, for your group label, and 
  * The (i) prediction; and (ii) classified maps

If you did **NOT** save these you're in a bit of trouble ...

---

## Question #1

* Construct maps showing the sample points, by your respective group and its unique objective


```{r}
# code for the question goes inside chunks, if part of the exercise solution.
#  If not part of the exercise questions delete this.
#  NOTE: do NOT answer all questions from within the same chunk
```

**YET ANOTHER NOTE RELATED TO CODE:**  
If King's English interpretation is required, it will go here, after each code chunk.

---

## Question #2

* Export field sample points as a shapefile

---

## Question #3

* Save your data as R objects:
  * All field campaign maps as R spatial object
  * All field campaign maps as **`.img`** format
* Save these R objects in a **`.RData`** file as well

---

## The End

---
