---
title: "EX06"
author: "Reed Scott"
date: "3/2/2022"
output: html_document
---

```{r}
#sdhmR Exercise 06
#02/23/2022
#RCS & LAP

#----------Global options---------#
#Setting Paths
path.root <- "D:/OneDrive - University of Vermont/Classes/Spring2022/sdhmR/sdhmR-V2022.1" #Reed Path
#path.root <- "C:/Users/14842/Documents/SDHM/sdhmR-V2022.1" #Lindsey Path
path.ex <- paste(path.root, "/data/exercise/traindat", sep = "") #Path to mod 2
path.preds <- paste(path.root, '/data/exercise/preds', sep = '')
path.figs <- paste(path.root, "/powerpoints/figures", sep = "") #path to save figs
path.gis <- paste(path.root, "/data/gis_layers", sep = "") #path to gis files

# some libraries
library(raster)	# FXNS: extent, raster, res, rasterize, values, writeRaster, 
#       cellFromXY, extract
library(sf)    	# FXNS: st_as_sf, st_as_sfc, st_bbox, st_convex_hull, st_union
#       st_buffer, st_read, st_write, st_intersection, st_geometry,
#       st_drop_geometry, st_coordinates, st_sf
library(terra)	# FXNS: as.data.frame, rast, vect
library(sp)
library(dplyr)
library(tidyverse)
library(data.table)
library(modeest)
# source course CRSs
setwd(path.root)
getwd()
source("r_code/prjs4_sdhmR.r")
ls(pattern = "prj.") # should be 5 prj.* objects

```

```{r}
######Question 1######
######################
# Break the predictor variables into logical groups of **topography**, **temperature**, and **precipitation**.  Do this by creating a new column called **`pred_type`** and add it the dataframe.

setwd(path.ex)
load('pers.trTOPO.RDATA')
table(pers.trTOPO['PERS106'])
  
unique(pers.trTOPO$PERS106)
unique(pers.trTOPO$etpt_5.x)
head(pers.trTOPO)

short <- pers.trTOPO[,27:42]
all_data <- short
pred_groups <- data.frame(names(all_data))
names(pred_groups) <- "Variable"
topo = c("exp1nrm.img",
         "exp3nrm.img",
         "exp5nrm.img",
         "prad_sw_di.img",
         "rough_1k.img",
         "topos.img")
temp = c("tave_s_hal.img",
         "tave_sprin.img",
         "tmax_s_hal.img",
         "tmax_summe.img")
precip = c("etpt_5.img",
           "etpt_6.img",
           "etpt_sprin.img",
           "mind_yr_av.img",
           "prec_w_hal.img",
           "prec_winte.img")
pred_groups$pred_type = if_else(pred_groups$Variable %in% topo,
                                "topography",
                                if_else(pred_groups$Variable %in% temp,
                                        "temperature",
                                        "precipitation"))

###########################################
```

```{r}
## Question #2
#* Assess correlations:
#  * Among all variables; and
#* Within each of the logical groupings created above

# numeric correlations
cut.point <- 0.7 # set cutpoint for correlation
c1 <- cor(pers.trTOPO[, c(27:42)],
          use = "pairwise.complete.obs", 
          method = "spearman") # est. correlation
c1 # examine
c2 <- subset(c1 > cut.point | c1 < -cut.point) # matrix of cor>cutpoint
c2 # examine; FALSE indicates cor<cutpoint Not correlated if false?

###Topo Correlations
#############
c3 <- cor(dplyr::select(all_data, one_of(topo)), 
          use = "pairwise.complete.obs",
          method = "spearman")
c_topo <- subset(c3 > cut.point | c3 < -cut.point)
c_topo

#####
# Temp correlations
c4 <- cor(dplyr::select(all_data, one_of(temp)), use = "pairwise.complete.obs", method = "spearman")
c_temp <- subset(c4 > cut.point | c4 < -cut.point)
c_temp

# Precip correlations
c5 <- cor(dplyr::select(all_data, one_of(precip)),
          use = "pairwise.complete.obs",
          method = "spearman")
c_precip <- subset(c5 > cut.point | c5 < -cut.point)
c_precip


###################################################################

panel.cor <- function(x, y,
                      digits=2,
                      prefix="",
                      cex.cor) 
{ usr <- par("usr"); on.exit(par(usr)) 
par(usr = c(0, 1, 0, 1)) 
r <- abs(cor(x,
             y,
             use = "pairwise.complete.obs")) 
txt <- format(c(r,
                0.123456789),
              digits = digits)[1] 
txt <- paste(prefix,
             txt,
             sep = "") 
if(missing(cex.cor)) cex <- 0.8/strwidth(txt) 

test <- cor.test(x,y) 
# borrowed from printCoefmat
Signif <- symnum(test$p.value,
                 corr = FALSE,
                 na = FALSE, 
                 cutpoints = c(0, 0.001, 0.01, 0.05, 0.1, 1),
                 symbols = c("***", "**", "*", ".", " ")) 
text(0.5,
     0.5,
     txt,
     cex = cex * r)
text(.8,
     .8,
     Signif,
     cex = cex,
     col = 2) 
}

pairs(pers.trTOPO[,
              c(27:42)],
      lower.panel = panel.smooth, 
      upper.panel = panel.cor,
      main = "Topo Variables") 

```


```{r}
## Question #3

#* Examine variable importance of each predictor variable as related to presence:absence using the process described in the Module.     

#### START function variable importance
varimp.glm <- function(tr.spp,
                       tr.var,
                       pres,
                       pf,
                       pl) {
  tmp.mat <- matrix(ncol = 2,
                    nrow = (pl - pf + 1))
  for (i in pf:pl) {
    # option: linear+quadratic; linear only
    tmp <- glm(tr.spp[, pres] ~ tr.var[, i] + 
                 I((tr.var[, i])^2),
               na.action = na.omit, 
               family = binomial)
    # linear only glm
    #tmp <- glm(tr.spp[, pres] ~ tr.var[, i], na.action = na.omit, family = binomial)
    tmp.mat[(i - pf + 1), 1] <- tmp$aic
    tmp.mat[(i - pf + 1), 2] <- (1 - (tmp$deviance/tmp$null.deviance))
  }
  return(tmp.mat)
} 
#### END function variable importance

# estimate VIP values => AIC & Adj deviance
tr.vip <- pers.trTOPO[, c(2, 27:42)] # keep only P/A & predictors
pres <- 1 # column for presence:absence
v.start <- 2 # column start predictor variables
v.stop <- ncol(tr.vip) # last column predictor variables
v.num <- v.stop - 1 # number predictor variables
dev.fit <- varimp.glm(tr.vip,
                      tr.vip,
                      pres,
                      v.start,
                      v.stop) # call VIP function
dev.fit # output matrix; col=1 AIC, col=2 Adj deviance 

# built basic barplot if desired
d.max <- ceiling(signif(max(dev.fit[, 2]), 2) * 10)/10 # max of y-axis
ylim.r <- range(0, d.max) # range y-axis
x.labs <- names(tr.vip[2:v.stop]) # x-axis labels
barplot(dev.fit[, 2],
        col = "darkgreen",
        ylim = ylim.r,
        main = "topo VIPs", 
        ylab = "adj.D2",
        names = x.labs,
        las = 2) # barplot
abline(h = 0) # add horizontal line
abline(mean(dev.fit[, 2]),
       0,
       lt = 3) # ref lines; dash=mean adj.dev 


dev.fit <- varimp.glm(tr.vip, tr.vip, pres, v.start, v.stop)
dev.fit <- data.frame(pred_groups, dev.fit)

################################

### Determine VIP with GLMs
varimp.glm <- function(tr.spp, tr.var, pres, pf, pl) {
  tmp.mat <- matrix(ncol = 2, nrow = (pl - pf + 1))
  for (i in pf:pl) {
    tmp <- glm(tr.spp[, pres] ~ tr.var[, i] + I((tr.var[, i])^2), na.action = na.omit,
               family = binomial)
    tmp.mat[(i - pf + 1), 1] <- tmp$aic
    tmp.mat[(i - pf + 1), 2] <- (1 - (tmp$deviance/tmp$null.deviance))
  }
  return(tmp.mat)
} 

tr.vip <- pers.trTOPO[, c(2, 27:42)]
pres <- 1
v.start <- 2
v.stop <- ncol(tr.vip)
v.num <- v.stop - 1
dev.fit <- varimp.glm(tr.vip, tr.vip, pres, v.start, v.stop)
dev.fit <- data.frame(pred_groups, dev.fit)
setnames(dev.fit, c("X1", "X2"), c("AIC", "AdjDeviance"))

dev.fit1 <- dev.fit[order(dev.fit$pred_type),]
dev.fit$Variable <- factor(dev.fit$Variable, 
                           levels = dev.fit$Variable[order(dev.fit$pred_type)])


# Plot AdjDeviance by group
```
```{r}
library(ggplot2)
plot <- ggplot()+
  geom_col(data = dev.fit, 
           aes(Variable, 
               AdjDeviance,
               fill = pred_type)) +
  scale_y_continuous(limits = c(0,0.3),
                     expand = c(0,0))+
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5,
                                   hjust=1),
        panel.background = element_blank(),
        axis.line.x = element_line(color = "black"),
        axis.line.y = element_line(color = "black"))

plot
# save plots
#setwd(path.figs)
#savePlot(filename = "mod2.7fig03.pdf", type = "pdf")
######## END VARIABLE IMPORTANCE AMONG TOPOGARPHIC PREDICTORS
################################################################################
```

## Question #4

#* Eliminate redundant variables with a goal of retaining 7-10 of the 16 available
#* Justify your decision(s) to keep / remove variables using bullets

 - We eliminated exp1nrm, exp3nrm, exp5nrm, topos, rough_1k, prad_sw_di, and prec_w_hal because of their low adjusted deviances. All other variables had higher deviances (above our cut-off) and were not highly correlated.

```{r}
remove <- c("exp1nrm.img",
           "exp3nrm.img",
           "exp5nrm.img",
           "topos.img",
           "rough_1k.img",
           "prad_sw_di.img",
           "prec_w_hal.img")


tr_PERS = dplyr::select(pers.trTOPO, !one_of(remove))

### Below cut-off line, get rid of
### Redundant variables don't cross mean
### Variables that cross mean have most explanatory value

```


#---
  
  ## Question #5
  
#  * Include the term **tr_MORT**, **tr_PERS**, **tr_RANG**, and **tr_SEED** somewhere in the final training data objects, depending on your data group.  Use of **tr_ByGroup** here means the data are now ready for model creation.
#* Save your data as R objects:
#  * Dataframe;
#* Point shapefile with geometry in R
#* Export as a point shapefile in ESRI format
#* Save these R objects in a **`.RData`** file

#These data will be used as we next begin the SDHM model constructions.
```{r}
tr_PERS_SF <- st_as_sf(tr_PERS,
                       coords = c("tr.wgs_x", "tr.wgs_y"),
                       crs = prj.wgs84,
                       remove = F)
save(tr_PERS,
     tr_PERS_SF,
     file = "tr_PERS.Rdata")

st_write(tr_PERS,
         dsn = ".",
         layer = "tr_PERS_SF",
         driver = "ESRI Shapefile",
         delete_layer = T,
         delete_dsn = T)


```

