---
title: "Exercise Title Goes Here"
author: "Your Name Goes Here"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

<!-- These are global options  
       set root directory here 
       NOTE mine is active; yours is commented out
       change the path to reflect your CPU,
       then comment out or delete mine
--> 
```{r global_options, include=FALSE}
knitr::opts_knit$set(root.dir = "~/words/classes/sdmR_ALLversions/sdhmR-V2022.1")
#knitr::opts_knit$set(root.dir = "~/sdhmR-V2022.1")
knitr::opts_chunk$set(warning=FALSE,error=TRUE,message=FALSE)
```

---

## This exercise links to Module 2.6

Submit the completed exercise to us at:

* t.edwards@usu.edu 
* eric_tsakiris@fws.gov  

---

## Context

The data are of the *Pinus edulis* Common piñon, the species you dealt with in Exercise #3 -- #5.  This exercise relies on the dataframe you generated in exercise #5.  Hopefully you saved those combined presence:absence dataframe, or regrettably you will need to return to exercise #5 and re--build that dataframe.

The goal is to examine the characteristics of the extracted predictor variable values as a part of the SDHM analytical process to possibly eliminate some variables from the models.

Remember, you will have been assigned data for *edulis* having one the following four discrete labels:

* **Seedlings**:  these data indicate spatial locations where seedlings of the tree have been found
* **Mortality**: these data indicate spatial locations where mortality has been observed
* **Persistence**: these data have no observed mortality or seedling at given spatial locations
* **Range**: the total dataset, including all spatial locations of seedlings, mortality, and persistence

---

## The Data

* The dataframe completed in exercise #5.

---

## Question #1

* Break the predictor variables into logical groups of **topography**, **temperature**, and **precipitation**.  Do this by creating a new column called **`pred_type`** and add it the dataframe.

```{r}
# code for the question goes inside chunks, if part of the exercise solution.
#  If not part of the exercise questions delete this.
#  NOTE: do NOT answer all questions from within the same chunk
```

**YET ANOTHER NOTE RELATED TO CODE:**  
If King's English interpretation is required, it will go here, after each code chunk.

---

## Question #2

* Assess correlations:
  * Among all variables; and
  * Within each of the logical groupings created above

---

## Question #3

* Examine variable importance of each predictor variable as related to presence:absence using the process described in the Module.

---

## Question #4

* Eliminate redundant variables with a goal of retaining 7-10 of the 16 available
* Justify your decision(s) to keep / remove variables using bullets

---

## Question #5

* Include the term **tr_MORT**, **tr_PERS**, **tr_RANG**, and **tr_SEED** somewhere in the final training data objects, depending on your data group.  Use of **tr_ByGroup** here means the data are now ready for model creation.
* Save your data as R objects:
  * Dataframe;
  * Point shapefile with geometry in R
  * Export as a point shapefile in ESRI format
* Save these R objects in a **`.RData`** file

These data will be used as we next begin the SDHM model constructions.

---

## The End

---
