---
title: "Exercise 8"
author: "Group J: Lindsey Pekurney and Reed Scott"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

<!-- These are global options  
       set root directory here 
       NOTE mine is active; yours is commented out
       change the path to reflect your CPU,
       then comment out or delete mine
--> 
```{r global_options, include=FALSE}
knitr::opts_knit$set(root.dir = "D:/OneDrive - University of Vermont/Classes/Spring2022/sdhmR/sdhmR-V2022.1") #Reed Path
#knitr::opts_knit$set(root.dir ="C:/Users/14842/Documents/SDHM/sdhmR-V2022.1") #Lindsey Path
#knitr::opts_knit$set(root.dir = "~/sdhmR-V2022.1")
knitr::opts_chunk$set(warning=FALSE,error=TRUE,message=FALSE)
```

```{r global option}
#----------Global options---------#
#Setting Paths
path.root <- "D:/OneDrive - University of Vermont/Classes/Spring2022/sdhmR/sdhmR-V2022.1" #Reed Laptop Path
#path.root <- "C:/Users/14842/Documents/SDHM/sdhmR-V2022.1" #Lindsey Path
path.ex <- paste(path.root,
                 "/data/exercise/traindat",
                 sep = "") #Path to mod 2
path.preds <- paste(path.root,
                    '/data/exercise/preds',
                    sep = '')
path.figs <- paste(path.root,
                   "/powerpoints/figures",
                   sep = "") #path to save figs
path.gis <- paste(path.root,
                  "/data/gis_layers",
                  sep = "") #path to gis files

# some libraries
library(raster)	# FXNS: extent, raster, res, rasterize, values, writeRaster, 
#       cellFromXY, extract
library(sf)    	# FXNS: st_as_sf, st_as_sfc, st_bbox, st_convex_hull, st_union
#       st_buffer, st_read, st_write, st_intersection, st_geometry,
#       st_drop_geometry, st_coordinates, st_sf
library(terra)	# FXNS: as.data.frame, rast, vect
library(sp)
library(dplyr)
library(tidyverse)
library(data.table)
library(modeest)
 library(PresenceAbsence)
  library(DAAG)
  library(car) 
  library(gam)
  options( warn = -1 )
# source course CRSs
setwd(path.root)
getwd()
source("r_code/prjs4_sdhmR.r")
ls(pattern = "prj.") # should be 5 prj.* objects
```
---

## This exercise links to Module 5.2

Submit the completed exercise to us at:

* t.edwards@usu.edu 
* eric_tsakiris@fws.gov  

---

## Context

The data are of the *Pinus edulis* Common piñon, the species you dealt with in Exercise #3 -- #5.  This exercise relies on the dataframe you generated in exercise #6.  Hopefully you saved those combined presence:absence dataframe, with the reduced variables you begin SDHM model construction with, or regrettably you will need to return to exercise #6 and re--build that dataframe.

The goal is to:
* Variables in final model, significance, and direction of relationship
  * List which variables are "smoothed" vs, those that are not
* Plots of the "smoothed" variables
* Calculate an estimate of model fit
* Build a table of accuracy metrics, and AUC plot
* Provide some bulleted interpretation points
* Plot map products of the:
  * (i) probability; and 
  * (ii) classified distribution model
  * (Save these two maps as **`.img`** files for use later in the course)


Remember, you will have been assigned data for *edulis* having one the following four discrete labels:

* **Seedlings**:  these data indicate spatial locations where seedlings of the tree have been found
* **Mortality**: these data indicate spatial locations where mortality has been observed
* **Persistence**: these data have no observed mortality or seedling at given spatial locations
* **Range**: the total dataset, including all spatial locations of seedlings, mortality, and persistence

---

## The Data

* The dataframe completed in exercise #6.

---

## Question #1

* Import and explore data
  * **NOTE**:  intent is not to reduce number of variables; that was completed in exercise #5.  Rather, calculate simple descriptive statistics (mean, sd, n) and boxplots


```{r}

############################
#Question 1: Import and explore the data
setwd(path.ex)
load('tr_PERS.Rdata')
class(tr_PERS)
head(tr_PERS)
drops <- c('wgs_xF',
           'wgs_yF',
           'cell.wgs_x',
           'cell.wgs_y',
           'exp5nrm',
           'exp3nrm',
           'rough_1k',
           'prad_sw_di',
           'etpt_5',
           'mind_yr_av',
           'prec_winte',
           'tave_sprin',
           'UNIQUEID.x',
           'UNIQUEID.y')
tr_PERS <- tr_PERS[ ,!(names(tr_PERS) %in% drops)]
tr_PERS <- na.omit(tr_PERS)
dim(tr_PERS)

#attempt to use for() loop
descSTAT <- {}
descSTAT <- as.data.frame(descSTAT)
for(i in 13:length(tr_PERS)){
  mean <- mean(tr_PERS[,i],
               na.rm = T) #mean returned NA
  NAS <- sum(is.na(tr_PERS[,i])) #checked for number of NAs
  SD <- sd(tr_PERS[,i],
           na.rm = T) #standard deviation
  N <- nrow(tr_PERS) - sum(is.na(tr_PERS[,i])) #check total N
  rname <- names(tr_PERS[,i])
  stats <- c(rname,
             mean,
             NAS,
             SD,
             N)
  descSTAT <- rbind(descSTAT,
                    stats)
}

#explore etpt5
mean(tr_PERS$etpt_5.img,
     na.rm = T) #mean returned NA
sum(is.na(tr_PERS$etpt_5.img)) #checked for number of NAs
sd(tr_PERS$etpt_5.img,
   na.rm = T) #standard deviation
nrow(tr_PERS) - sum(is.na(tr_PERS$etpt_5.img)) #check total N

#explore etpt6
mean(tr_PERS$etpt_6.img,
     na.rm = T) #mean returned NA
sum(is.na(tr_PERS$etpt_6.img)) #checked for number of NAs
sd(tr_PERS$etpt_6.img,
   na.rm = T) #standard deviation
nrow(tr_PERS) - sum(is.na(tr_PERS$etpt_6.img)) #check total N

#explore etpt_sprin
mean(tr_PERS$etpt_sprin.img, na.rm = T) #mean returned NA
sum(is.na(tr_PERS$etpt_sprin.img)) #checked for number of NAs
sd(tr_PERS$etpt_sprin.img, na.rm = T) #standard deviation
nrow(tr_PERS) - sum(is.na(tr_PERS$etpt_sprin.img)) #check total N

#explore mind_yr_ave
mean(tr_PERS$mind_yr_av.img, na.rm = T) #mean returned NA
sum(is.na(tr_PERS$mind_yr_av.img)) #checked for number of NAs
sd(tr_PERS$mind_yr_av.img, na.rm = T) #standard deviation
nrow(tr_PERS) - sum(is.na(tr_PERS$mind_yr_av.img)) #check total N

#explore prec_winte
mean(tr_PERS$prec_winte.img, na.rm = T) #mean returned NA
sum(is.na(tr_PERS$prec_winte.img)) #checked for number of NAs
sd(tr_PERS$prec_winte.img, na.rm = T) #standard deviation
nrow(tr_PERS) - sum(is.na(tr_PERS$prec_winte.img)) #check total N

#explore tave_s_hal
mean(tr_PERS$tave_s_hal.img, na.rm = T) #mean returned NA
sum(is.na(tr_PERS$tave_s_hal.img)) #checked for number of NAs
sd(tr_PERS$tave_s_hal.img, na.rm = T) #standard deviation
nrow(tr_PERS) - sum(is.na(tr_PERS$tave_s_hal.img)) #check total N

#explore tave_sprin
mean(tr_PERS$tave_sprin.img, na.rm = T) #mean returned NA
sum(is.na(tr_PERS$tave_sprin.img)) #checked for number of NAs
sd(tr_PERS$tave_sprin.img, na.rm = T) #standard deviation
nrow(tr_PERS) - sum(is.na(tr_PERS$tave_sprin.img)) #check total N

#explore tmax_summe
mean(tr_PERS$tmax_summe.img, na.rm = T) #mean returned NA
sum(is.na(tr_PERS$tmax_summe.img)) #checked for number of NAs
sd(tr_PERS$tmax_summe.img, na.rm = T) #standard deviation
nrow(tr_PERS) - sum(is.na(tr_PERS$tmax_summe.img)) #check total N
```

```{r}
#Boxplots
par(mfrow = c(3, 3))
boxplot(tr_PERS$etpt_5.img ~ tr_PERS$PERS106, xlab = "Presence:Absence", 
        ylab = "etpt5", na.rm = T)
boxplot(tr_PERS$etpt_6.img ~ tr_PERS$PERS106, xlab = "Presence:Absence", 
        ylab = "etpt6", na.rm = T)
boxplot(tr_PERS$etpt_sprin.img ~ tr_PERS$PERS106, xlab = "Presence:Absence", 
        ylab = "etpt_sprin", na.rm = T)
boxplot(tr_PERS$mind_yr_av.img ~ tr_PERS$PERS106, xlab = "Presence:Absence", 
        ylab = "mind_yr_ave", na.rm = T)
boxplot(tr_PERS$prec_winte.img ~ tr_PERS$PERS106, xlab = "Presence:Absence", 
        ylab = "prec_winte", na.rm = T)
boxplot(tr_PERS$tave_s_hal.img ~ tr_PERS$PERS106, xlab = "Presence:Absence", 
        ylab = "tave_s_hal", na.rm = T)
boxplot(tr_PERS$tave_sprin.img ~ tr_PERS$PERS106, xlab = "Presence:Absence", 
        ylab = "tave_sprin", na.rm = T)
boxplot(tr_PERS$tmax_s_hal.img ~ tr_PERS$PERS106, xlab = "Presence:Absence", 
        ylab = "tmax_s_hal", na.rm = T)
boxplot(tr_PERS$tmax_summe.img ~ tr_PERS$PERS106, xlab = "Presence:Absence", 
        ylab = "tmax_summe", na.rm = T)
```


---

## Question #2

* Construct some basic plots to assess possible relationships warranting use of a GAM
* List which variables may be amenable to a GAM--based smoother

I believe all variables may be amenable to a GAM-based smoother, but I think etpt5, etpt6, etptsprin, and precwinte will definitely require smooters.
```{r}

plot(tr_PERS[1:13]) # simple data scatterplots
# fancy scatterplot w/lines if desired from library(car)
  scatterplotMatrix(tr_PERS[1:13],
                    main = "Scatterplots of RS data")  # scatterplot

# factor plots
  par(mfcol = c(1, 3)) # if desired
  col.list <- c("grey80",
                "forestgreen") # specified color ramp
  
  plot(as.factor(tr_PERS$PERS106) ~ tr_PERS$etpt_5.img,
       xlab = "ETPT_5",
       ylab = "Persistence", 
    col = col.list)
  
  plot(as.factor(tr_PERS$PERS106) ~ tr_PERS$etpt_6.img,
       xlab = "ETPT_6", 
    ylab = "Persistence",
    col = col.list)
  
  plot(as.factor(tr_PERS$PERS106) ~ tr_PERS$etpt_sprin.img,
       xlab = "ETPT_Sprin",
       ylab = "Persistence", 
    col = col.list)
  
   plot(as.factor(tr_PERS$PERS106) ~ tr_PERS$prec_winte.img,
       xlab = "Prec Winte",
       ylab = "Persistence", 
    col = col.list)
  
    plot(as.factor(tr_PERS$PERS106) ~ tr_PERS$tave_s_hal.img,
       xlab = "Tave S Hal",
       ylab = "Persistence", 
    col = col.list)
  
    plot(as.factor(tr_PERS$PERS106) ~ tr_PERS$tave_sprin.img,
       xlab = "Tave Sprin",
       ylab = "Persistence", 
    col = col.list)
    
    plot(as.factor(tr_PERS$PERS106) ~ tr_PERS$tmax_s_hal.img,
       xlab = "Tmax S Hal",
       ylab = "Persistence", 
    col = col.list)
    
    plot(as.factor(tr_PERS$PERS106) ~ tr_PERS$tmax_summe.img,
       xlab = "Tmax Summe",
       ylab = "Persistence", 
    col = col.list)
  


```



---

## Question #3

* Construct a set of GAMs from fully linear to a maximum of 5 splines
  * Use lowess smoother rather than default spline
```{r}
################################################################################
######## START COMPARATIVE GAM MODELLING

# GAM model 0: all linear
#   can use glm for linear model 
  mod0.GLM <- glm(PERS106 ~ etpt_5.img +
                    etpt_6.img +
                    etpt_sprin.img +
                    mind_yr_av.img +
                    prec_winte.img +
                    tave_s_hal.img +
                    tave_sprin.img +
                    tmax_s_hal.img +
                    tmax_summe.img, 
    family = binomial,
    data = tr_PERS)

# alternative linear model using gam call
  mod0.GAM <- gam(PERS106 ~ etpt_5.img +
                    etpt_6.img +
                    etpt_sprin.img +
                    mind_yr_av.img +
                    prec_winte.img +
                    tave_s_hal.img +
                    tave_sprin.img +
                    tmax_s_hal.img +
                    tmax_summe.img, 
    family = binomial,
    data = tr_PERS,
    )
  summary(mod0.GAM) # model #0 summary

# GAM model 1: all smoothers using defaults df (smoothers=4)
  mod1.GAM <- gam(PERS106 ~ s(etpt_5.img) +
                    s(etpt_6.img) +
                    s(etpt_sprin.img) +
                    s(mind_yr_av.img) +
                    s(prec_winte.img) +
                    s(tave_s_hal.img) +
                    s(tave_sprin.img) +
                    s(tmax_s_hal.img) +
                    s(tmax_summe.img), 
    family = binomial,
    data = tr_PERS)
  summary(mod1.GAM) # model #1 summary

# GAM model 2: all smoothers using specified df (smoothers=5)
  mod2.GAM <- gam(PERS106 ~ s(etpt_5.img, 5) +
                    s(etpt_6.img, 5) +
                    s(etpt_sprin.img, 5) +
                    s(mind_yr_av.img, 5) +
                    s(prec_winte.img, 5) +
                    s(tave_s_hal.img, 5) +
                    s(tave_sprin.img, 5) +
                    s(tmax_s_hal.img, 5) +
                    s(tmax_summe.img, 5), 
                  family = binomial, 
                  data = tr_PERS)
  summary(mod2.GAM) # model #2 summary 

# GAM model 3: stepwise w/diff scopes
  mod3.GAM <- step.Gam(mod2.GAM,
                       scope = list("etpt_5.img" = ~1 + 
                                    etpt_5.img + 
                                    s(etpt_5.img, 3) + 
                                    s(etpt_5.img, 5),
                                  "etpt_6.img" = ~1 + 
                                    etpt_6.img + 
                                    s(etpt_6.img, 3) + 
                                    s(etpt_6.img, 5),
                                  "etpt_sprin.img" = ~1 + 
                                    etpt_sprin.img + 
                                    s(etpt_sprin.img, 3) + 
                                    s(etpt_sprin.img, 5),
                                  "mind_yr_av.img" = ~1 + 
                                    mind_yr_av.img + 
                                    s(mind_yr_av.img, 3) + 
                                    s(mind_yr_av.img, 5),
                                  "prec_winte.img" = ~ 1 + 
                                    prec_winte.img + 
                                    s(prec_winte.img, 3) + 
                                    s(prec_winte.img, 5),
                                  "tave_s_hal.img" = ~1 + 
                                    tave_s_hal.img + 
                                    s(tave_s_hal.img, 3) + 
                                    s(tave_s_hal.img, 5),
                                  "tave_sprin.img" = ~1 + 
                                    tave_sprin.img + 
                                    s(tave_sprin.img, 3) + 
                                    s(tave_sprin.img, 5),
                                  "tmax_s_hal.img" = ~1 + 
                                    tmax_s_hal.img + 
                                    s(tmax_s_hal.img, 3) + 
                                    s(tmax_s_hal.img, 5),
                                  "tmax_summe.img" = ~1 + 
                                    tmax_summe.img + 
                                    s(tmax_summe.img, 3) + 
                                    s(tmax_summe.img, 5)), 
                                  trace = F)
  summary(mod3.GAM) # model #3 summary
  
  
  names(mod3.GAM$coefficients) # find the linear terms
  #summary.glm(mod3.GAM)$coefficients[c(1:2, 4:6, 10:11), ] # access linear terms
######## END MODEL COMPARATIVE GAM MODELLING
################################################################################



```

---

## Question #4

* Calculate accuracy metrics (as in Module 3.2 and 3.3, Analytical Intermission) using:
  * Resubstitution approaches, and
  * A 10-fold cross--validation approach
```{r}
######## START MODEL COMPARISONS
# estimate model fits
  mod0.glm <- 100 * (1 - mod0.GLM$deviance/mod0.GLM$null.deviance) # model fit
  mod0.fit <- 100 * (1 - mod0.GAM$deviance/mod0.GAM$null.deviance) # model fit
  mod1.fit <- 100 * (1 - mod1.GAM$deviance/mod1.GAM$null.deviance) # model fit
  mod2.fit <- 100 * (1 - mod2.GAM$deviance/mod2.GAM$null.deviance) # model fit
  mod3.fit <- 100 * (1 - mod3.GAM$deviance/mod3.GAM$null.deviance) # model fit
  mods.fit <- cbind(mod0.glm, mod0.fit, mod1.fit, mod2.fit, mod3.fit) # all model fits
  mods.fit # examine 

# measures of resubstitution accuracy
  library(PresenceAbsence) # load pkg for accuracy
  mod.list <- c("mod0.GAM",
                "mod1.GAM",
                "mod2.GAM",
                "mod3.GAM")
  for (i in 1:length(mod.list)) {
    pred <- predict(get(mod.list[i]),
                    type = "response") # predict by model
    mod.num <- mod.list[i] # assign model No. to varname
    dat <- cbind(mod.num,
                 tr_PERS[2],
                 pred) # build obs and prediction dataframe
    cut <- optimal.thresholds(dat,
                              opt.methods = c("PredPrev=Obs")) # threshold=prevalence
    cfmat <- table(dat[[2]],
                   factor(as.numeric(dat$pred >= cut$pred))) # confusion matrix
    acc <- presence.absence.accuracy(dat,
                                     threshold = cut$pred,
                                     st.dev = F) # calc accuracies
    tss <- acc$sensitivity + acc$specificity - 1 # code TSS metric
    acc <- cbind(acc[1:7], tss) # bind all metrics
    acc$model <- mod.num # variable substitution
    if (i == 1) mods.acc <- acc else mods.acc <- rbind(mods.acc, acc) # build output dataframe
    }
  mods.acc[c(1, 4:5, 7:8)] # resubstittution accuracies by model 


# measures of 5-fold cross-validation accuracy
#   absent set.seed() answers will differ among runs
  library(DAAG) # load pkg for crossval
  set.seed(1234) # set seed for replicability
  mod.list <- c("mod0.GAM",
                "mod1.GAM",
                "mod2.GAM",
                "mod3.GAM")
  for (i in 1:length(mod.list)) {
    cv.mod <- CVbinary(get(mod.list[i]),
                       nfolds = 10,
                       print.details = F) # crossval predict
    cv.pers <- cv.mod$cvhat # assign new name to jackknife estimates
    mod.num <- mod.list[i] # assign model No. to varname
    cv.dat <- cbind(mod.num,
                    tr_PERS[2],
                    cv.pers) # build obs and prediction dataframe
    cv.cut <- optimal.thresholds(cv.dat,
                                 opt.methods = c("ObsPrev")) # threshold=pers
    cv.cfmat <- table(cv.dat[[2]],
                      factor(as.numeric(cv.dat$cv.pers > cv.cut$cv.pers))) # confusion matrix
    cv.acc <- presence.absence.accuracy(cv.dat,
                                        threshold = cv.cut$cv.pers,
                                        st.dev = F) # calculate accuracies
    cv.tss <- cv.acc$sensitivity + cv.acc$specificity - 1 # code TSS metric
    cv.acc <- cbind(cv.acc[1:7],
                    cv.tss) # bind all metrics
    cv.acc$model <- mod.num # variable substitution
    if (i == 1) mods.cvacc <- cv.acc else mods.cvacc <- rbind(mods.cvacc, cv.acc) # build output dataframe
    }
  mods.cvacc[c(1, 4:5, 7:8)]  # accuracies by model 
######## END MODEL COMPARISONS
```

---

## Question #5

* Build 2 prediction maps:
  * A raw probability estimate for each cell in the modelling domain; and
  * A classified map based on the selected threshold from Question #4

```{r}
setwd(path.preds)
pers.list <- list.files(pattern = ".img$") # list of .img files; $ strips extra
pers.list # examine
pers.dom <- stack(pers.list) # build raster stack
pers.dom # examine stack
names(pers.dom) # examine stack names
names(pers.dom) <- paste(names(pers.dom),'img')

setwd(path.gis)
states <- st_read(dsn = ".",
                  layer = "na_states_wgs") # import shapefile



#create a probability model
pers.prob <- predict(pers.dom,
                     mod1.GAM, 
                     type = "response",
                     fun = predict,
                     index = 2,
                     overwrite = T) # prediction raster
pers.prob # examine



# next reclassify based on threshold mod.cut per above
pers.class <- reclassify(pers.prob,
                         c(0,cut[[2]],0,
                           cut[[2]],1,1),
                         overwrite=TRUE)
pers.class

#restrict class and prob models to domain
setwd(path.ex)
load('pers.PPsA.RData')
pers.bufR <-raster(pers.bufR)
new.pers.bufR <- projectRaster(pers.bufR,
                               pers.class)
pers.class <- pers.class*new.pers.bufR
pers.prob <- pers.prob*new.pers.bufR
#Plot probability map#
par(mfrow = c(1,1))
plot(pers.prob,
     axes = T,
     main = 'Probability Map')
plot(st_geometry(states),
     add = T,
     lwd = 1.5)

#Classification Map#
plot(pers.class, legend = F,
     axes = T,
     main = "Classification Map") # plot classification map
plot(st_geometry(states),
     add = T,
     lwd = 1.5) # add state boundaries 


```

---

## Question #6

* Save your data as R objects:
  * Accuracy metrics as a dataframe;
  * Classification threshold as a stand--alone scalar
  * Both prediction maps as **`.img`** format
* Save these R objects in a **`.RData`** file

These data will be used again in Module 10, Ensemble Models.
```{r}
setwd('path.ex')
save.image(pers.class,
           pers.prob,
           file = 'ex8')
save(cut,
     mods.acc,
     pers.class,
     pers.prob, 
     file = 'ex8.RData')
save.image()
```

```
---

## The End

---
