---
title: "Exercise Title Goes Here"
author: "Your Name Goes Here"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

<!-- These are global options  
       set root directory here 
       NOTE mine is active; yours is commented out
       change the path to reflect your CPU,
       then comment out or delete mine
--> 
```{r global_options, include=FALSE}
knitr::opts_knit$set(root.dir = "~/words/classes/sdmR_ALLversions/sdhmR-V2022.1")
#knitr::opts_knit$set(root.dir = "~/sdhmR-V2022.1")
knitr::opts_chunk$set(warning=FALSE,error=TRUE,message=FALSE)
```

---

## This exercise links to Module 5.2

Submit the completed exercise to us at:

* t.edwards@usu.edu 
* eric_tsakiris@fws.gov  

---

## Context

The data are of the *Pinus edulis* Common piñon, the species you dealt with in Exercise #3 -- #5.  This exercise relies on the dataframe you generated in exercise #6.  Hopefully you saved those combined presence:absence dataframe, with the reduced variables you begin SDHM model construction with, or regrettably you will need to return to exercise #6 and re--build that dataframe.

The goal is to:
* Variables in final model, significance, and direction of relationship
  * List which variables are "smoothed" vs, those that are not
* Plots of the "smoothed" variables
* Calculate an estimate of model fit
* Build a table of accuracy metrics, and AUC plot
* Provide some bulleted interpretation points
* Plot map products of the:
  * (i) probability; and 
  * (ii) classified distribution model
  * (Save these two maps as **`.img`** files for use later in the course)


Remember, you will have been assigned data for *edulis* having one the following four discrete labels:

* **Seedlings**:  these data indicate spatial locations where seedlings of the tree have been found
* **Mortality**: these data indicate spatial locations where mortality has been observed
* **Persistence**: these data have no observed mortality or seedling at given spatial locations
* **Range**: the total dataset, including all spatial locations of seedlings, mortality, and persistence

---

## The Data

* The dataframe completed in exercise #6.

---

## Question #1

* Import and explore data
  * **NOTE**:  intent is not to reduce number of variables; that was completed in exercise #5.  Rather, calculate simple descriptive statistics (mean, sd, n) and boxplots


```{r}
# code for the question goes inside chunks, if part of the exercise solution.
#  If not part of the exercise questions delete this.
#  NOTE: do NOT answer all questions from within the same chunk
```

**YET ANOTHER NOTE RELATED TO CODE:**  
If King's English interpretation is required, it will go here, after each code chunk.

---

## Question #2

* Construct some basic plots to assess possible relationships warranting use of a GAM
* List which variables may be amenable to a GAM--based smoother

---

## Question #3

* Construct a set of GAMs from fully linear to a maximum of 5 splines
  * Use lowess smoother rather than default spline

---

## Question #4

* Calculate accuracy metrics (as in Module 3.2 and 3.3, Analytical Intermission) using:
  * Resubstitution approaches, and
  * A 10-fold cross--validation approach

---

## Question #5

* Build 2 prediction maps:
  * A raw probability estimate for each cell in the modelling domain; and
  * A classified map based on the selected threshold from Question #4

---

## Question #6

* Save your data as R objects:
  * Accuracy metrics as a dataframe;
  * Classification threshold as a stand--alone scalar
  * Both prediction maps as **`.img`** format
* Save these R objects in a **`.RData`** file

These data will be used again in Module 10, Ensemble Models.

---

## The End

---
