---
title: "Exercise Title Goes Here"
author: "Your Name Goes Here"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

<!-- These are global options  
       set root directory here 
       NOTE mine is active; yours is commented out
       change the path to reflect your CPU,
       then comment out or delete mine
--> 
```{r global_options, include=FALSE}
knitr::opts_knit$set(root.dir = "C:/Users/14842/Documents/SDHM/sdhmR-V2022.1/SDHM_Reed/ex8")
#knitr::opts_knit$set(root.dir = "~/sdhmR-V2022.1")
knitr::opts_chunk$set(warning=FALSE,
                      error= FALSE,
                      message=FALSE)
```

```{r global option}
#----------Global options---------#
#Setting Paths
#path.root <- "D:/OneDrive - University of Vermont/Classes/Spring2022/sdhmR/sdhmR-V2022.1" #Reed Laptop Path
path.root <- "C:/Users/14842/Documents/SDHM/sdhmR-V2022.1" #Lindsey Path
path.ex <- paste(path.root,
                 "/data/exercise/traindat",
                 sep = "") #Path to mod 2
path.preds <- paste(path.root,
                    '/data/exercise/preds',
                    sep = '')
path.figs <- paste(path.root,
                   "/powerpoints/figures",
                   sep = "") #path to save figs
path.gis <- paste(path.root,
                  "/data/gis_layers",
                  sep = "") #path to gis files

# some libraries
library(raster)	# FXNS: extent, raster, res, rasterize, values, writeRaster, 
#       cellFromXY, extract
library(sf)    	# FXNS: st_as_sf, st_as_sfc, st_bbox, st_convex_hull, st_union
#       st_buffer, st_read, st_write, st_intersection, st_geometry,
#       st_drop_geometry, st_coordinates, st_sf
library(terra)	# FXNS: as.data.frame, rast, vect
library(sp)
library(dplyr)
library(tidyverse)
library(data.table)
library(modeest)
 library(PresenceAbsence)
  library(DAAG)
  library(car) 
  library(gam)
  options( warn = -1 )
# source course CRSs
setwd(path.root)
getwd()
source("C:/Users/14842/Documents/SDHM/sdhmR-V2022.1/r_code/prjs4_sdhmR.r")
ls(pattern = "prj.") # should be 5 prj.* objects
```
---

## This exercise links to Module 5.2

Submit the completed exercise to us at:

* t.edwards@usu.edu 
* eric_tsakiris@fws.gov  

---

## Context

The data are of the *Pinus edulis* Common piñon, the species you dealt with in Exercise #3 -- #5.  This exercise relies on the dataframe you generated in exercise #6.  Hopefully you saved those combined presence:absence dataframe, with the reduced variables you begin SDHM model construction with, or regrettably you will need to return to exercise #6 and re--build that dataframe.

The goal is to:
* Variables in final model, significance, and direction of relationship
  * List which variables are "smoothed" vs, those that are not
* Plots of the "smoothed" variables
* Calculate an estimate of model fit
* Build a table of accuracy metrics, and AUC plot
* Provide some bulleted interpretation points
* Plot map products of the:
  * (i) probability; and 
  * (ii) classified distribution model
  * (Save these two maps as **`.img`** files for use later in the course)


Remember, you will have been assigned data for *edulis* having one the following four discrete labels:

* **Seedlings**:  these data indicate spatial locations where seedlings of the tree have been found
* **Mortality**: these data indicate spatial locations where mortality has been observed
* **Persistence**: these data have no observed mortality or seedling at given spatial locations
* **Range**: the total dataset, including all spatial locations of seedlings, mortality, and persistence

---

## The Data

* The dataframe completed in exercise #6.

---

## Question #1

* Import and explore data
  * **NOTE**:  intent is not to reduce number of variables; that was completed in exercise #5.  Rather, calculate simple descriptive statistics (mean, sd, n) and boxplots


```{r}

############################
#Question 1: Import and explore the data

load('C:/Users/14842/Documents/SDHM/sdhmR-V2022.1/SDHM_Reed/ex6/tr_PERS.Rdata')
class(tr_PERS)
head(tr_PERS)
drops <- c('wgs_xF',
           'wgs_yF',
           'cell.wgs_x',
           'cell.wgs_y',
           'exp5nrm',
           'exp3nrm',
           'rough_1k',
           'prad_sw_di',
           'etpt_5',
           'mind_yr_av',
           'prec_winte',
           'tave_sprin',
           'UNIQUEID.x',
           'UNIQUEID.y')
tr_PERS <- tr_PERS[ ,!(names(tr_PERS) %in% drops)]
tr_PERS <- na.omit(tr_PERS)
dim(tr_PERS)

#attempt to use for() loop
descSTAT <- {}
descSTAT <- as.data.frame(descSTAT)
for(i in 13:length(tr_PERS)){
  mean <- mean(tr_PERS[,i],
               na.rm = T) #mean returned NA
  NAS <- sum(is.na(tr_PERS[,i])) #checked for number of NAs
  SD <- sd(tr_PERS[,i],
           na.rm = T) #standard deviation
  N <- nrow(tr_PERS) - sum(is.na(tr_PERS[,i])) #check total N
  rname <- names(tr_PERS[,i])
  stats <- c(rname,
             mean,
             NAS,
             SD,
             N)
  descSTAT <- rbind(descSTAT,
                    stats)
}

#explore etpt5
mean(tr_PERS$etpt_5.img,
     na.rm = T) #mean returned NA
sum(is.na(tr_PERS$etpt_5.img)) #checked for number of NAs
sd(tr_PERS$etpt_5.img,
   na.rm = T) #standard deviation
nrow(tr_PERS) - sum(is.na(tr_PERS$etpt_5.img)) #check total N

#explore etpt6
mean(tr_PERS$etpt_6.img, na.rm = T) #mean returned NA
sum(is.na(tr_PERS$etpt_6.img)) #checked for number of NAs
sd(tr_PERS$etpt_6.img, na.rm = T) #standard deviation
nrow(tr_PERS) - sum(is.na(tr_PERS$etpt_6.img)) #check total N

#explore etpt_sprin
mean(tr_PERS$etpt_sprin.img, na.rm = T) #mean returned NA
sum(is.na(tr_PERS$etpt_sprin.img)) #checked for number of NAs
sd(tr_PERS$etpt_sprin.img, na.rm = T) #standard deviation
nrow(tr_PERS) - sum(is.na(tr_PERS$etpt_sprin.img)) #check total N

#explore mind_yr_ave
mean(tr_PERS$mind_yr_av.img, na.rm = T) #mean returned NA
sum(is.na(tr_PERS$mind_yr_av.img)) #checked for number of NAs
sd(tr_PERS$mind_yr_av.img, na.rm = T) #standard deviation
nrow(tr_PERS) - sum(is.na(tr_PERS$mind_yr_av.img)) #check total N

#explore prec_winte
mean(tr_PERS$prec_winte.img, na.rm = T) #mean returned NA
sum(is.na(tr_PERS$prec_winte.img)) #checked for number of NAs
sd(tr_PERS$prec_winte.img, na.rm = T) #standard deviation
nrow(tr_PERS) - sum(is.na(tr_PERS$prec_winte.img)) #check total N

#explore tave_s_hal
mean(tr_PERS$tave_s_hal.img, na.rm = T) #mean returned NA
sum(is.na(tr_PERS$tave_s_hal.img)) #checked for number of NAs
sd(tr_PERS$tave_s_hal.img, na.rm = T) #standard deviation
nrow(tr_PERS) - sum(is.na(tr_PERS$tave_s_hal.img)) #check total N

#explore tave_sprin
mean(tr_PERS$tave_sprin.img, na.rm = T) #mean returned NA
sum(is.na(tr_PERS$tave_sprin.img)) #checked for number of NAs
sd(tr_PERS$tave_sprin.img, na.rm = T) #standard deviation
nrow(tr_PERS) - sum(is.na(tr_PERS$tave_sprin.img)) #check total N

#explore tmax_summe
mean(tr_PERS$tmax_summe.img, na.rm = T) #mean returned NA
sum(is.na(tr_PERS$tmax_summe.img)) #checked for number of NAs
sd(tr_PERS$tmax_summe.img, na.rm = T) #standard deviation
nrow(tr_PERS) - sum(is.na(tr_PERS$tmax_summe.img)) #check total N
```

```{r}
#Boxplots
par(mfrow = c(3, 3))
boxplot(tr_PERS$etpt_5.img ~ tr_PERS$PERS106, xlab = "Presence:Absence", 
        ylab = "etpt5", na.rm = T)
boxplot(tr_PERS$etpt_6.img ~ tr_PERS$PERS106, xlab = "Presence:Absence", 
        ylab = "etpt6", na.rm = T)
boxplot(tr_PERS$etpt_sprin.img ~ tr_PERS$PERS106, xlab = "Presence:Absence", 
        ylab = "etpt_sprin", na.rm = T)
boxplot(tr_PERS$mind_yr_av.img ~ tr_PERS$PERS106, xlab = "Presence:Absence", 
        ylab = "mind_yr_ave", na.rm = T)
boxplot(tr_PERS$prec_winte.img ~ tr_PERS$PERS106, xlab = "Presence:Absence", 
        ylab = "prec_winte", na.rm = T)
boxplot(tr_PERS$tave_s_hal.img ~ tr_PERS$PERS106, xlab = "Presence:Absence", 
        ylab = "tave_s_hal", na.rm = T)
boxplot(tr_PERS$tave_sprin.img ~ tr_PERS$PERS106, xlab = "Presence:Absence", 
        ylab = "tave_sprin", na.rm = T)
boxplot(tr_PERS$tmax_s_hal.img ~ tr_PERS$PERS106, xlab = "Presence:Absence", 
        ylab = "tmax_s_hal", na.rm = T)
boxplot(tr_PERS$tmax_summe.img ~ tr_PERS$PERS106, xlab = "Presence:Absence", 
        ylab = "tmax_summe", na.rm = T)
```


---

## Question #2

* Construct some basic plots to assess possible relationships warranting use of a GAM
* List which variables may be amenable to a GAM--based smoother

```{r}
######## START SOME PRELIMINARY FIGURES
# some scatterplot of remote sensing-based variables
  plot(tr_PERS[20:27]) # simple data scatterplots
# fancy scatterplot w/lines if desired from library(car)
  scatterplotMatrix(tr_PERS[20:27],
                    main = "Scatterplots of RS data")  # scatterplot

# factor plots
  par(mfcol = c(1, 3)) # if desired
  col.list <- c("grey80",
                "forestgreen") # specified color ramp
  
  plot(as.factor(tr_PERS$PERS106) ~ tr_PERS$etpt_5.img,
       xlab = "ETPT_5",
       ylab = "Persistence", 
    col = col.list) # plot 1: tm4
  
  plot(as.factor(dat1$lp) ~ dat1$precip,
       xlab = "precipitation", 
    ylab = "Lodgeple category",
    col = col.list) # plot 2: precipitation
  
  plot(as.factor(dat1$lp) ~ dat1$elev,
       xlab = "Elevation",
       ylab = "Lodgeple category", 
    col = col.list) # plot 3: elevation
  par(mfrow = c(1, 1))

```
---

## Question #3

* Construct a set of GAMs from fully linear to a maximum of 5 splines
  * Use lowess smoother rather than default spline

---

## Question #4

* Calculate accuracy metrics (as in Module 3.2 and 3.3, Analytical Intermission) using:
  * Resubstitution approaches, and
  * A 10-fold cross--validation approach

---

## Question #5

* Build 2 prediction maps:
  * A raw probability estimate for each cell in the modelling domain; and
  * A classified map based on the selected threshold from Question #4

---

## Question #6

* Save your data as R objects:
  * Accuracy metrics as a dataframe;
  * Classification threshold as a stand--alone scalar
  * Both prediction maps as **`.img`** format
* Save these R objects in a **`.RData`** file

These data will be used again in Module 10, Ensemble Models.

---

## The End

---
