---
title: "Exercise 11"
author: "Group J"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

<!-- These are global options  
       set root directory here 
       NOTE mine is active; yours is commented out
       change the path to reflect your CPU,
       then comment out or delete mine
--> 
```{r global_options, include=FALSE}
knitr::opts_knit$set(root.dir = "D:/OneDrive - University of Vermont/Classes/Spring2022/sdhmR/sdhmR-V2022.1") #Reed Path
#knitr::opts_knit$set(root.dir ="C:/Users/14842/Documents/SDHM/sdhmR-V2022.1") #Lindsey Path
#knitr::opts_knit$set(root.dir = "~/sdhmR-V2022.1")
knitr::opts_chunk$set(warning=FALSE,error=TRUE,message=FALSE)
```

---

## This exercise links to Module 5.5

Submit the completed exercise to us at:

* t.edwards@usu.edu 
* eric_tsakiris@fws.gov  

---

## Context

The data are of the *Pinus edulis* Common piñon, the species you dealt with in Exercise #3 -- #5.  This exercise relies on the dataframe you generated in exercise #6.  Hopefully you saved those combined presence:absence dataframe, with the reduced variables you begin SDHM model construction with, or regrettably you will need to return to exercise #6 and re--build that dataframe.

The goal is to:
* Produce and interpret variable importance plots
* Build a table of accuracy metrics, and AUC plot
* Plot map products of the:
  * (i) probability; and 
  * (ii) classified distribution model
  * (Save these two maps as **`.img`** files for use later in the course)


Remember, you will have been assigned data for *edulis* having one the following four discrete labels:

* **Seedlings**:  these data indicate spatial locations where seedlings of the tree have been found
* **Mortality**: these data indicate spatial locations where mortality has been observed
* **Persistence**: these data have no observed mortality or seedling at given spatial locations
* **Range**: the total dataset, including all spatial locations of seedlings, mortality, and persistence

---

## The Data

* The dataframe completed in exercise #6.

---

## Question #1

* Import and explore data
  * **NOTE**:  intent is not to reduce number of variables; that was completed in exercise #5.  Rather, calculate simple descriptive statistics (mean, sd, n) and boxplots


```{r Creat paths + load libs}
#Initialize -----------------------------------------
path.root <- "D:/OneDrive - University of Vermont/Classes/Spring2022/sdhmR/sdhmR-V2022.1" #Reed Laptop Path
#path.root <- "C:/Users/14842/Documents/SDHM/sdhmR-V2022.1" #Lindsey Path
path.ex <- paste(path.root, "/data/exercise/traindat", sep = "") #Path to mod 2
path.preds <- paste(path.root, '/data/exercise/preds', sep = '')
path.figs <- paste(path.root, "/powerpoints/figures", sep = "") #path to save figs
path.gis <- paste(path.root, "/data/gis_layers", sep = "") #path to gis files

# some libraries
library(raster)	# FXNS: extent, raster, res, rasterize, values, writeRaster, 
#       cellFromXY, extract
library(sf)    	# FXNS: st_as_sf, st_as_sfc, st_bbox, st_convex_hull, st_union
#       st_buffer, st_read, st_write, st_intersection, st_geometry,
#       st_drop_geometry, st_coordinates, st_sf
library(terra)	# FXNS: as.data.frame, rast, vect
library(sp)
library(dplyr)
library(tidyverse)
library(data.table)
library(modeest)
library(DAAG)
library(PresenceAbsence)
options("rgdal_show_exportToProj4_warnings"="none") # run this before library(rgdal)
library(rgdal)
library(dismo)
library(maptools)
library(randomForest)

# source course CRSs
setwd(path.root)
getwd()
source("r_code/prjs4_sdhmR.r")
ls(pattern = "prj.") # should be 5 prj.* objects
```


```{r import data}
# Global Variables-------------------------------------
setwd(path.ex)
load('tr_PERS.RData')
class(tr_PERS)
head(tr_PERS)
drops <- c('wgs_xF','wgs_yF','cell.wgs_x','cell.wgs_y','exp5nrm','exp3nrm','rough_1k','prad_sw_di',
           'etpt_5','mind_yr_av','prec_winte','tave_sprin','UNIQUEID.x'
           ,'UNIQUEID.y')
tr_PERS <- tr_PERS[ ,!(names(tr_PERS) %in% drops)]
tr_PERS <- na.omit(tr_PERS)
dim(tr_PERS)
```

```{r explore data}
# Program Body------------------------------------------

###Question 1: explore data:###
#explore etpt5
mean(tr_PERS$etpt_5.img, na.rm = T) #mean returned NA
sum(is.na(tr_PERS$etpt_5.img)) #checked for number of NAs
sd(tr_PERS$etpt_5.img, na.rm = T) #standard deviation
nrow(tr_PERS) - sum(is.na(tr_PERS$etpt_5.img)) #check total N

#explore etpt6
mean(tr_PERS$etpt_6.img, na.rm = T) #mean returned NA
sum(is.na(tr_PERS$etpt_6.img)) #checked for number of NAs
sd(tr_PERS$etpt_6.img, na.rm = T) #standard deviation
nrow(tr_PERS) - sum(is.na(tr_PERS$etpt_6.img)) #check total N
#explore etpt_sprin
mean(tr_PERS$etpt_sprin.img, na.rm = T) #mean returned NA
sum(is.na(tr_PERS$etpt_sprin.img)) #checked for number of NAs
sd(tr_PERS$etpt_sprin.img, na.rm = T) #standard deviation
nrow(tr_PERS) - sum(is.na(tr_PERS$etpt_sprin.img)) #check total N
#explore mind_yr_ave
mean(tr_PERS$mind_yr_av.img, na.rm = T) #mean returned NA
sum(is.na(tr_PERS$mind_yr_av.img)) #checked for number of NAs
sd(tr_PERS$mind_yr_av.img, na.rm = T) #standard deviation
nrow(tr_PERS) - sum(is.na(tr_PERS$mind_yr_av.img)) #check total N
#explore prec_winte
mean(tr_PERS$prec_winte.img, na.rm = T) #mean returned NA
sum(is.na(tr_PERS$prec_winte.img)) #checked for number of NAs
sd(tr_PERS$prec_winte.img, na.rm = T) #standard deviation
nrow(tr_PERS) - sum(is.na(tr_PERS$prec_winte.img)) #check total N
#explore tave_s_hal
mean(tr_PERS$tave_s_hal.img, na.rm = T) #mean returned NA
sum(is.na(tr_PERS$tave_s_hal.img)) #checked for number of NAs
sd(tr_PERS$tave_s_hal.img, na.rm = T) #standard deviation
nrow(tr_PERS) - sum(is.na(tr_PERS$tave_s_hal.img)) #check total N
#explore tave_sprin
mean(tr_PERS$tave_sprin.img, na.rm = T) #mean returned NA
sum(is.na(tr_PERS$tave_sprin.img)) #checked for number of NAs
sd(tr_PERS$tave_sprin.img, na.rm = T) #standard deviation
nrow(tr_PERS) - sum(is.na(tr_PERS$tave_sprin.img)) #check total N
#explore tmax_summe
mean(tr_PERS$tmax_summe.img, na.rm = T) #mean returned NA
sum(is.na(tr_PERS$tmax_summe.img)) #checked for number of NAs
sd(tr_PERS$tmax_summe.img, na.rm = T) #standard deviation
nrow(tr_PERS) - sum(is.na(tr_PERS$tmax_summe.img)) #check total N


#Boxplots
par(mfrow = c(3, 3))
boxplot(tr_PERS$etpt_5.img ~ tr_PERS$PERS106, xlab = "Presence:Absence", 
        ylab = "etpt5", na.rm = T)
boxplot(tr_PERS$etpt_6.img ~ tr_PERS$PERS106, xlab = "Presence:Absence", 
        ylab = "etpt6", na.rm = T)
boxplot(tr_PERS$etpt_sprin.img ~ tr_PERS$PERS106, xlab = "Presence:Absence", 
        ylab = "etpt_sprin", na.rm = T)
boxplot(tr_PERS$mind_yr_av.img ~ tr_PERS$PERS106, xlab = "Presence:Absence", 
        ylab = "mind_yr_ave", na.rm = T)
boxplot(tr_PERS$prec_winte.img ~ tr_PERS$PERS106, xlab = "Presence:Absence", 
        ylab = "prec_winte", na.rm = T)
boxplot(tr_PERS$tave_s_hal.img ~ tr_PERS$PERS106, xlab = "Presence:Absence", 
        ylab = "tave_s_hal", na.rm = T)
boxplot(tr_PERS$tave_sprin.img ~ tr_PERS$PERS106, xlab = "Presence:Absence", 
        ylab = "tave_sprin", na.rm = T)
boxplot(tr_PERS$tmax_s_hal.img ~ tr_PERS$PERS106, xlab = "Presence:Absence", 
        ylab = "tmax_s_hal", na.rm = T)
boxplot(tr_PERS$tmax_summe.img ~ tr_PERS$PERS106, xlab = "Presence:Absence", 
        ylab = "tmax_summe", na.rm = T)
par(mfrow=c(1,1))
```

**YET ANOTHER NOTE RELATED TO CODE:**  
If King's English interpretation is required, it will go here, after each code chunk.

---

## Question #2

* Construct a boosted regression tree (BRT) model using the presence:absence data
* Build plots of variable importance values

```{r Create model}
resp <- paste("as.factor(", colnames(tr_PERS[2]), ")", sep = "") # assign resp to col number
n.col <- ncol(tr_PERS) # number of columns
pred <- 5:n.col # assign predictors to column numbers 

################################################################################
######## START BRT MODELS
date() # start time stamp
pers.BRT <- gbm.step(data = tr_PERS, gbm.x = pred, gbm.y = 2, family = "bernoulli",
                     tree.complexity = 3, learning.rate = 1e-04, bag.fraction = 0.75, n.folds = 10, 
                     n.trees = 50, plot.main = TRUE, keep.fold.fit = TRUE) 
date()

#run again#
pers.BRT2 <- gbm.step(data = tr_PERS, gbm.x = pred, gbm.y = 2, family = "bernoulli",
                     tree.complexity = 3, learning.rate = 0.01, bag.fraction = 0.75, n.folds = 10, 
                     n.trees = 50, plot.main = TRUE, keep.fold.fit = TRUE) 
date()
```
---

## Question #3

* Calculate accuracy metrics (as in Module 3.2 and 3.3, Analytical Intermission) using:
  * Resubstitution approaches, and
  * A 10-fold cross--validation approach

---

## Question #4

* Build 2 prediction maps:
  * A raw probability estimate for each cell in the modelling domain; and
  * A classified map based on the selected threshold from Question #3
setwd(path.gis)
states <- st_read(dsn = ".", layer = "na_states_wgs") # import shapefile
```{r Create Models}
#create a probability model
setwd(path.preds)
load('pers.topoDOM.RData')
pers.dom <- pers.topoDOM
pers.prob.BRT <- predict(pers.dom, pers.BRT2, 
                        type = "response", n.trees = pers.BRT2$gbm.call$best.trees) # prediction raster
pers.prob.BRT # examine 

# next reclassify based on threshold mod.cut per above
pers.class.BRT <- reclassify(pers.prob.BRT, c(0,mod.cut[[2]],0, 
                                            mod.cut[[2]],1,1),overwrite=TRUE)
pers.class.BRT
```

```{r Restrict to domain}
#restrict to domain#
setwd(path.ex)
load("pers.bufptR.img")
# # crop(x = pers.topoDOM, y = )
#load('pers.PPsA.RData',verbose = TRUE)
# plot(pers.topoDOM)
# pers.bufR <-raster(pers.bufR)
# plot(pers.bufR)

new.pers.bufptR <- raster::projectRaster(pers.bufptR, pers.class.BRT)

pers.class.BRT <- pers.class.BRT*new.pers.bufptR
pers.prob.BRT <- pers.prob.BRT*new.pers.bufptR
```

```{r Plot maps}
par(mfrow=c(1,1))
#Plot probability map#
plot(pers.prob.BRT, axes = T, main = 'Probability Map')
plot(st_geometry(states), add = T)

#Classification Map#
plot(pers.class.BRT, legend = F, axes = T, main = "Classification Map") # plot classification map
plot(st_geometry(states), add = T, lwd = 1.5) # add state boundaries
```
---

## Question #5

* Save your data as R objects:
  * Accuracy metrics as a dataframe;
  * Classification threshold as a stand--alone scalar
  * Both prediction maps as **`.img`** format
* Save these R objects in a **`.RData`** file

These data will be used again in Module 10, Ensemble Models.
```{r Save}
setwd(path.ex)
save(pers.BRT,pers.BRT2, file = 'ex11.RData')
save(pers.class.BRT, file = 'persClassBRT.img')
save(pers.prob.BRT, file = 'persProbBRT.img')
```

---

## The End

---
